---
title: "Main"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

```{r include = FALSE}
library(knitr)
```

# SuperCAT
SUPERvised single Cell Annotation Tool

### Introduction

This document contains code workflows, explaining how to perform the cell annotation procedures for the single cell analyses described in ***A single-cell transcriptomic analysis of endometriosis, endometriomas, eutopic endometrial samples and uninvolved ovary tissues highlights cell populations characteristic of these tissue types. Transcriptional and cellular heterogeneity across tissues suggests novel therapeutic targets and biomarkers for this disease.***

### Authors


Marcos A.S. Fonseca\*, Marcela Haro\*, Kelly N. Wright\*, Xianzhi Lin\*, Forough Abbasi, Jennifer Sun, Lourdes Hernandez, Natasha L. Orr, Jooyoon Hong, Yunhee Choi-Kuaea, Horacio M. Maluf, Bonnie L. Balzer, Aaron Fishburn, Ryan Hickey, Ilana Cass, Helen S. Goodridge, Mireille Truong, Yemin Wang, Margareta D. Pisarska, Huy Dinh, Amal EL-Naggar, David Huntsman, Michael S. Anglesio, Marc T. Goodman, Fabiola Medeiros†, Matthew Siedhoff†, Kate Lawrenson†



#### Workflow for cell annotation

Please navigate to the document tabs above to acess the procedures in `step 1` and `step 2` for the single cell cluster annotations

``` {r message=F, echo=FALSE, warning=F, fig.align='center', out.width = "70%"}
img1_path = "img/annot_workflow.pdf-1.png"
include_graphics(img1_path)

```


#### Document authors
Marcos A. S. Fonseca


