---
title: "SuperCAT - Step 1"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })

---


```{r include = FALSE}
rm(list = ls())

library(ggplot2)
library(ComplexHeatmap)
library(circlize)
library(Seurat)
library(dplyr)
library(RColorBrewer)
library (pheatmap)
library(scales)

aux.seurat = readRDS("../AtlasEndometriosis/rds/aux.seurat.rds")

```

### Introduction

The procedures in this step aim to determine the enrichment of cell type markers in each cell cluster. Thus, the overall assumption is that the cells relationship in each group also determine the cell type specificity that characterize each cluster.

#### Pre-requirements

To start the procedures we consider:

* set of Seurat cell clusters with a high-resolution parameter (2 - 3).
* list of cell type markers.

For this tutorial the following clusters were considered

```{r include = T, eval=T, fig.align = 'center', fig.width = 10, fig.height=6, fig.align='center'}
DimPlot(aux.seurat, reduction="umap")
```


#### Differential expression of genes per cluster

The differential expression analysis is performed using MAST approach. It may take few hours depending on the number of cells and clusters.

``` {r message=F, eval=FALSE, warning=F, fig.align='center', out.width = "80%"}
markers <- FindAllMarkers(aux.seurat, test.use = "MAST")
write.table(markers, file = "DEG_seurat_clusters_res3.csv", quote = F, sep = ",")
```

#### Table of cell type marker genes

``` {r message=F, warning=F, fig.align='center', out.width = "80%"}
table.gene.markers <- data.frame(Gene=c("EPCAM", "KRT8", "KRT18", "KRT19", "KRT7", "FOXJ1"), Cell="Epithelial")
table.gene.markers <- rbind(table.gene.markers, data.frame(Gene=c("ACTA2"), Cell="Smooth muscle"))
table.gene.markers <- rbind(table.gene.markers, data.frame(Gene=c("DCN", "COL11A2", "FAP", "PDGFRA", "COL11A1", "COL1A1", "PDGFRB"), Cell="Fibroblast"))
table.gene.markers <- rbind(table.gene.markers, data.frame(Gene=c("LYZ", "CD14", "MME", "C1QA", "CLEC10A"), Cell="Myeloid"))
table.gene.markers <- rbind(table.gene.markers, data.frame(Gene=c("CLDN5", "PECAM1", "CD34", "ESAM"), Cell="Endothelial"))
table.gene.markers <- rbind(table.gene.markers, data.frame(Gene=c("CD79A"), Cell="Plasma"))
table.gene.markers <- rbind(table.gene.markers, data.frame(Gene=c("JCHAIN"), Cell="B Cells"))
table.gene.markers <- rbind(table.gene.markers, data.frame(Gene=c("TYROBP", "FCGR3A"), Cell="Natural Killer"))
table.gene.markers <- rbind(table.gene.markers, data.frame(Gene=c("TPSB2"), Cell="Mast"))
table.gene.markers <- rbind(table.gene.markers, data.frame(Gene=c("HBB", "GYPA"), Cell="Erythrocytes"))
table.gene.markers <- rbind(table.gene.markers, data.frame(Gene=c("CD2", "CD3D", "CD3E", "CD3G", 'CD8A', "CCL5", "PTPRC"), Cell="T Cells"))

grid::grid.newpage()
gridExtra::grid.table(table.gene.markers, rows=NULL)

```

Auxiliary functions to count the number of cell marker genes that are differentialy expressed in each cluster considering fold change >= 0.2 and p-value <= 0.05


``` {r message=F, warning=F}

markers = read.delim("files/DEG_seurat_clusters_res3.csv", sep = ",")

get_markers <- function(cell) {
  return(table.gene.markers$Gene[which(table.gene.markers$Cell == cell)])
}

count_marker_DEG <- function(table.gene.markers, deg, fc=0.2, pval=0.05) {

  clusters = unique(deg$cluster)
  counts = list()

  for (x in (unique(table.gene.markers$Cell))) {
    table = NULL
    for (i in clusters) {
      filt = deg[which(deg$cluster == i),]
      if (nrow(filt) > 0) {
        res = sum(filt$avg_log2FC >= fc & filt$p_val_adj <= pval & filt$gene %in% get_markers(cell = x))
      }else
        res = 0

      table = rbind(table, data.frame(Cluster=i, MarkesDEG=res))
    }
    counts[[x]] <- table
  }
  return(counts)
}

```

Lets now count the DEG marks from our initial list.

``` {r message=F, warning=F}

counts_deg = count_marker_DEG(table.gene.markers, markers, fc = 0.1)
counts_deg = do.call(cbind, counts_deg)
rownames(counts_deg) = counts_deg$Epithelial.Cluster

```

Lets normalize the counts by the number of the total corresponding markers. Visualizing the enrichment of marks in a heatmap

``` {r message=F, warning=F, fig.align='center', fig.height=17, fig.width=8}

row.bar = data.frame(table(Idents(aux.seurat)))
row_ha = rowAnnotation(ClusterSize = anno_barplot(row.bar$Freq))

markers.count = data.frame(table(table.gene.markers$Cell))
markers.count = markers.count[match(colnames(counts_deg[,seq(2, ncol(counts_deg), by = 2)]), paste0(markers.count$Var1, ".MarkesDEG")),]
markers.count$ColumnName = paste0(markers.count$Var1, " (n=", markers.count$Freq, ")")
head(markers.count)

y = counts_deg[,seq(2, ncol(counts_deg), by = 2)]
x = markers.count$Freq
z = (t(t(y) / x))

rownames(z) = rownames(counts_deg)
colnames(z) = markers.count$ColumnName
head(z)

ht1 = Heatmap(z,
              na_col = "white",
              cluster_rows = T,
              cluster_columns = F,
              show_row_names = T,
              row_names_gp = gpar(fontsize = 12),
              row_names_side = "left",
              rect_gp = gpar(col = "black", lwd = 0.6),
              show_column_names = T,
              right_annotation = row_ha,
              col = colorRamp2(c(0, mean(as.matrix(z)), max(z)), c("#FFFFFF" , "#d8b365", "#01665e"))

)

draw(ht1, heatmap_legend_side = "right")
```


Next, the expression of the markers can also be evaluated using Seurat dotplot function. We first change the order of the rows to match the order of the markers count clustering, obtained previously.

``` {r message=F, warning=F, fig.align='center', fig.height=17, fig.width=9}

aux.seurat@active.ident <- factor(aux.seurat@active.ident,
                                  levels=rownames(z)[rev(row_order(ht1))]
)

DotPlot(object = aux.seurat, features = rev(unique(table.gene.markers$Gene))) +
  scale_colour_gradient2(low = "#1b7837", mid = "#f7f7f7", high = "#762a83") +
  RotatedAxis() +
  theme(axis.text.x = element_text(size = 12))

```

### Cluster assignments procedures

In this session we defined some rules to sistematiquely assign the cell type

``` {r message=F, warning=F, fig.align='center'}
expr.markers = DotPlot(object = aux.seurat, features = rev(unique(table.gene.markers$Gene)))

expr.markers.data = expr.markers$data
krt.genes = table.gene.markers$Gene[grep("KRT", table.gene.markers$Gene)]

my_cluster_col <- data.frame(row.names = rownames(z))
my_cluster_col$CellType = NA

for (r in 1:nrow(z)) {

  c = r -1
  nmarkers = which(z[r,] > 0)
  pct.KRT = expr.markers.data[expr.markers.data$id == c & expr.markers.data$features.plot %in% krt.genes,]

  if ( (sum(pct.KRT$pct.exp > 35) > 0 &  sum(pct.KRT$avg.exp.scaled > 1) > 0) ) {
    my_cluster_col$CellType[r] = "Epithelial"
  } else if ( length(nmarkers) > 1  ){

    nmarkers = z[r,which(z[r,] > 0)]

    if ( length(grep("Smooth muscle", names(nmarkers))) > 0 ){

      pos.fibro = grep("Fibroblast", names(nmarkers))
      if ( length(pos.fibro) > 0 ) {
        if (nmarkers[pos.fibro] > 0.25) {
          my_cluster_col$CellType[r] = "Fibroblast"
        }else {
          my_cluster_col$CellType[r] = "Smooth"
        }
      } else {
        marker = which.max(nmarkers)
        my_cluster_col$CellType[r] = gsub(" .*", "", names(marker))
      }

    } else {
      nmarkers = z[r,which(z[r,] > 0)] * markers.count$Freq[which(markers.count$ColumnName %in% names(nmarkers))]

      if (length(unique(nmarkers)) == 1) {
        nmarkers = z[r,which(z[r,] > 0)]
        marker = which.max(nmarkers)
      } else {
        marker = which.max(nmarkers)
        print("skip")
      }
      my_cluster_col$CellType[r] = gsub(" .*", "", names(marker))
    }

  } else if (length(nmarkers) == 1) {
    my_cluster_col$CellType[r] = gsub(" .*", "", names(nmarkers))
  }

}

my_cluster_col

write.table(my_cluster_col, 
            "files/reference_panel_clusters_assigned.cvs", 
            sep = ",",
            quote = FALSE)

```
