---
title: "SuperCAT - Step 2"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })

---


```{r include = FALSE}
rm(list = ls())

library(ComplexHeatmap)
library(circlize)
library(Seurat)
#library(ggplot2)
#library(ggrepel)
library(dplyr)
#library(plotly)
#library(cowplot)
#library(tidyr)
library(corrplot)
library(RColorBrewer)
library (pheatmap)

aux.seurat = readRDS("../AtlasEndometriosis/rds/aux.seurat.rds")

```

### Introduction

The procedures in this step aim to determine the enrichment of cell type markers in each cell cluster. Thus, the overall assumption is that the cells relationship in each group also determine the cell type specificity that characterize each cluster.

#### Pre-requirements

After calculating the DEG per cluster, lets take a look into the profile signature up to the top 100 genes per cluster.

``` {r message=F, warning=F, fig.align='center', fig.height=8, fig.width=10}
markers = read.delim("files/DEG_seurat_clusters_res3.csv", sep = ",")

ng = 100
p.v = 0.05
fc = 0
features.deg <- markers %>%
  filter(p_val_adj < p.v & avg_log2FC > fc) %>%
  group_by(cluster) %>%
  top_n(ng)

aux = DotPlot(aux.seurat, features = unique(features.deg$gene))
aux.data = aux$data

data.avg.exp.scaled = data.frame(row.names = unique(features.deg$gene))

for (o in unique(aux.data$id) ) {
  aux.data.filt = aux.data[which(aux.data$id == o),]
  rownames(aux.data.filt) <- aux.data.filt$features.plot
  data.avg.exp.scaled = cbind(data.avg.exp.scaled, aux.data.filt[, 'avg.exp.scaled', drop=FALSE])
}

colnames(data.avg.exp.scaled) <- paste(unique(aux.data$id), " (n=", table(features.deg$cluster), ")", sep = "")
pheatmap(data.avg.exp.scaled, 
         show_rownames = F, 
         main = paste0("DEG expression signature clusters (n=", length(unique(features.deg$gene)), ")"), 
         cutree_cols = 10 )

```


``` {r message=F, warning=F, fig.align='center',  fig.height=8, fig.width=10}

my_cluster_col = read.delim("files/reference_panel_clusters_assigned.cvs", sep = ",")

my_cluster_col$cor.val = NA
my_cluster_col$cor.clus = NA

cor.mat = cor(data.avg.exp.scaled)
colnames(cor.mat) <- paste(0:113, sep = "")

for (r in 1:nrow(my_cluster_col)) {

  c = r -1
  if (is.na(my_cluster_col$CellType[r]))  {
    print(c)
    cor.mat.filt = cor.mat[r,]
    cor.mat.filt = cor.mat.filt[-r]

    my_cluster_col$cor.val[r] = max(cor.mat.filt)
    pos.clus = as.numeric(names(which.max(cor.mat.filt)))
    my_cluster_col$cor.clus[r] = pos.clus

    my_cluster_col$CellType[r] = my_cluster_col$CellType[(pos.clus + 1)]
  }
}

pheatmap(cor.mat,
         annotation_col = my_cluster_col)


```

``` {r message=F, eval=FALSE, warning=F, fig.align='center', out.width = "80%"}
aux.seurat@active.ident <- plyr::mapvalues(x = aux.seurat@active.ident, from = rownames(my_cluster_col), to = my_cluster_col$CellType)
aux.seurat[["active.cluster"]] <- aux.seurat@active.ident
cluster.cols = c("#9e0142", "#abd9e9", "#abdda4", "#d53e4f", "#f46d43", "#5e4fa2", "#fdae61", "#ffffbf", "#e6f598",  "#66c2a5", "#3288bd","#6DCD59FF","#35B779FF","#1F9E89FF","#26828EFF","#31688EFF","#3E4A89FF","#482878FF","#440154FF")

DimPlot(object = aux.seurat, pt.size = 0.1, cols = cluster.cols, label = T)
```

``` {r message=F, eval=TRUE, echo=F, warning=F, fig.align='center', out.width = "80%"}
cluster.cols = c("#9e0142", "#abd9e9", "#abdda4", "#d53e4f", "#f46d43", "#5e4fa2", "#fdae61", "#ffffbf", "#e6f598",  "#66c2a5", "#3288bd","#6DCD59FF","#35B779FF","#1F9E89FF","#26828EFF","#31688EFF","#3E4A89FF","#482878FF","#440154FF")

DimPlot(object = aux.seurat, pt.size = 0.1, cols = cluster.cols, label = T, group.by = "active.cluster")
```